---
title: "Distribution maps Unionidae"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE)
opts_knit$set(root.dir = here())

# Conflicting packages
conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)
conflicted::conflicts_prefer(dplyr::lag)

# Packages
library(tidyverse)
library(ggpattern)
library(sf)
library(readxl)

# Source
source(here("source", "read_excel_allsheets.R"))
```

# Read in data
## Map layers

We read in different map layers to create distribution maps.

```{r, results='hide'}
proj_crs <- 32631

# Read in map layers
belgium_sf <- st_read(here("data", "maps", "Belgie.shp")) %>%
  st_transform(proj_crs)
utm5_total_sf <- st_read(here("data", "maps", "utm5.shp")) %>%
  st_transform(proj_crs)
utm10_total_sf <- st_read(here("data", "maps", "utm10.shp")) %>%
  st_transform(proj_crs)
rivieren_sf <- st_read(here("data", "maps", "hoofdrivierenbel.shp")) %>%
  st_transform(proj_crs)
kanalen_sf <- st_read(here("data", "maps", "Kanalen.shp")) %>%
  st_transform(proj_crs)
```

```{r}
# Select UTM grids that overlap with Belgium
umt5_tags <- utm5_total_sf %>%
  st_intersection(belgium_sf) %>%
  pull(TAG)

utm5_sf <- utm5_total_sf %>%
  filter(TAG %in% umt5_tags)

umt10_tags <- utm10_total_sf %>%
  st_intersection(belgium_sf) %>%
  pull(TAG)

utm10_sf <- utm10_total_sf %>%
  filter(TAG %in% umt10_tags)

# Combine water ways Belgium
waterlopen_sf <- bind_rows(
  rivieren_sf %>%
    select("naam" = NAAM) %>%
    mutate(waterloop = "rivier"),
  kanalen_sf %>%
    select("naam" = NAAM) %>%
    mutate(waterloop = "kanaal"))
```

We visualise the layers.

```{r}
ggplot() +
  geom_sf(data = belgium_sf, fill = "white",
          linewidth = 0.8) +
  geom_sf(data = waterlopen_sf, aes(colour = waterloop),
          linewidth = 0.8) +
  scale_color_manual(values = c("cornflowerblue", "darkblue")) +
  geom_sf(data = utm5_sf, fill = alpha("white", 0),
          linewidth = 0.3, colour = "grey") +
  geom_sf(data = utm10_sf, fill = alpha("white", 0),
          linewidth = 0.6) +
  labs(title = "UTM 5 km", x = "", y = "", colour = "Waterloop") +
  coord_sf(datum = proj_crs) +
  theme_void() +
  theme(legend.position = c(0.2, 0.2))

ggplot() +
  geom_sf(data = belgium_sf, fill = "white",
          linewidth = 0.8) +
  geom_sf(data = waterlopen_sf, aes(colour = waterloop),
          linewidth = 0.8) +
  scale_color_manual(values = c("cornflowerblue", "darkblue")) +
  geom_sf(data = utm10_sf, fill = alpha("white", 0),
          linewidth = 0.6) +
  labs(title = "UTM 10 km", x = "", y = "", colour = "Waterloop") +
  coord_sf(datum = proj_crs) +
  theme_void() +
  theme(legend.position = c(0.2, 0.2))
```

## Occurrence data

We read in and prepare the data from after 1995.
Taxonomic information was added manually.

```{r}
# Read in data post 1995
data_succinea_raw <- read_excel_allsheets(
  here("data", "raw",
       "AA-Locaties-Najaden-2023-10-25-vanaf1995-Final.xlsx")
  )

# Get species list
species_list <- lapply(names(data_succinea_raw), function(sheet) {
  data_succinea_raw[[sheet]][1,] %>%
    select(Genus, Species, Author) %>%
    mutate(name = sheet)
})
species_df <- do.call(rbind.data.frame, species_list)

# Prepare data post 1995
data_succinea <- lapply(names(data_succinea_raw), function(sheet) {
  data_succinea_raw[[sheet]] %>%
    select(Locality:Alive, -c(`Lat., Long.`, "Date")) %>%
    mutate(name = sheet) %>%
    rename_with(~tolower(gsub("-", "_", .x))) %>%
    rename_with(~gsub( " .*$", "", .x))
})

data_post_1995 <- do.call(rbind.data.frame, data_succinea) %>%
  full_join(species_df, by = join_by(name)) %>%
  select(-name) %>%
  rename_with(tolower) %>%
  pivot_longer(cols = c("empty", "alive"),
               names_to = "state",
               values_to = "number") %>%
  drop_na(number) %>%
  rowid_to_column(var = "id") %>%
  mutate(period = "post_1995",
         id = paste(id, period, sep = "_"))

glimpse(data_post_1995)
```

We calculate UTM 5 km grid squares based on coordinates.

```{r}
data_post_1995_sf_raw <- data_post_1995 %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(proj_crs)

# Only select data within Belgium
data_post_1995_sf <- data_post_1995_sf_raw %>%
  st_intersection(belgium_sf) %>%
  select(-names(belgium_sf))

# Get UTM 5 km tags
intersection_occurrences_utm5 <- st_intersection(data_post_1995_sf, utm5_sf) %>%
  st_drop_geometry() %>%
  select(id, TAG)

# Get UTM 10 km tags
intersection_occurrences_utm10 <- st_intersection(data_post_1995_sf, utm10_sf) %>%
  st_drop_geometry() %>%
  select(id, TAG)

data_post_1995_sf <- data_post_1995_sf %>%
  full_join(intersection_occurrences_utm5, by = join_by(id)) %>%
  rename(utm_5km = TAG) %>%
  full_join(intersection_occurrences_utm10, by = join_by(id)) %>%
  rename(utm_10km_new = TAG) %>%
  mutate(test_tag = substr(utm_5km, 1, 4),
         test = ifelse(utm_10km == test_tag, TRUE, FALSE))
```

Are there mismatches between the UTM designation in the datasets and recalculated UTM tags (10 km)?

```{r}
mismatch_utm <- data_post_1995_sf %>%
  st_drop_geometry() %>%
  filter(test == FALSE) %>%
  select(id, utm_5km, utm_10km_new) %>%
  left_join(data_post_1995, by = join_by(id)) %>%
  select(id, locality, genus, species, author, state, longitude, latitude,
         utm_10km, utm_5km_new = utm_5km, utm_10km_new)
mismatch_utm %>%
  kable()
```

Are there locations that fall outside Belgium?

```{r}
outside_be <- data_post_1995_sf_raw %>%
  st_difference(belgium_sf) %>%
  st_drop_geometry() %>%
  select(id) %>%
  left_join(data_post_1995, by = join_by(id)) %>%
  select(id, locality, genus, species, author, state, longitude, latitude,
         utm_10km)

outside_be %>%
  kable()
```

We read in and prepare the data from before 1995.

```{r}
# Read in data pre 1995
data_nijs_raw <- read_excel_allsheets(
  here("data", "raw",
       "Alfa-Unionidae 6 soorten-Verspreiding-Nijs-1995-Empty-Alive.xlsx")
)

# Prepare data pre 1995
data_nijs <- lapply(names(data_nijs_raw), function(sheet) {
  data_nijs_raw[[sheet]] %>%
    mutate(name = sheet) %>%
    rename("empty" = `Empty t/m 1994`, "alive" = `Alive t/m 1994`)
})

data_pre_1995 <- do.call(rbind.data.frame, data_nijs) %>%
  full_join(species_df, by = join_by(name)) %>%
  select(-name) %>%
  rename_with(tolower) %>%
  pivot_longer(cols = c("empty", "alive"),
               names_to = "state",
               values_to = "utm_10km") %>%
  rowid_to_column(var = "id") %>%
  mutate(period = "pre_1995",
         id = paste(id, period, sep = "_"))

glimpse(data_pre_1995)
```


# Map data

We create a dataset where each row is an utm grid square.

-  5 km post 1995
-  10 km post 1995
-  10 km pre 1995

```{r}
data_post_1995_utm5km <- data_post_1995_sf %>%
  st_drop_geometry() %>%
  distinct(genus, species, author, state, period, utm_5km, utm_10km_new) %>%
  mutate(status = ifelse(state == "alive", "levend", "leeg")) %>%
  select(-state) %>%
  pivot_wider(id_cols = everything(), names_from = "status",
              values_from = "status") %>%
  mutate(state_fill = paste(levend, leeg, sep = " + "),
         state_fill = gsub("^NA\\s\\+\\s", "", state_fill),
         state_fill = gsub("\\s\\+\\sNA$", "", state_fill)) %>%
  mutate(spec_name = paste(genus, species, author)) %>%
  select(utm_5km, utm_10km = utm_10km_new, period, spec_name, genus, species,
         author, state_fill)

data_post_1995_utm10km <- data_post_1995_sf %>%
  st_drop_geometry() %>%
  distinct(genus, species, author, state, period, utm_10km_new) %>%
  mutate(status = ifelse(state == "alive", "levend", "leeg")) %>%
  select(-state) %>%
  pivot_wider(id_cols = everything(), names_from = "status",
              values_from = "status") %>%
  mutate(state_fill = paste(levend, leeg, sep = " + "),
         state_fill = gsub("^NA\\s\\+\\s", "", state_fill),
         state_fill = gsub("\\s\\+\\sNA$", "", state_fill)) %>%
  mutate(spec_name = paste(genus, species, author)) %>%
  select(utm_10km = utm_10km_new, period, spec_name, genus, species, author,
         state_fill)

data_pre_1995_utm10km <- data_pre_1995 %>%
  distinct(genus, species, author, state, period, utm_10km) %>%
  mutate(status = ifelse(state == "alive", "levend", "leeg")) %>%
  select(-state) %>%
  pivot_wider(id_cols = everything(), names_from = "status",
              values_from = "status") %>%
  mutate(state_fill = paste(levend, leeg, sep = " + "),
         state_fill = gsub("^NA\\s\\+\\s", "", state_fill),
         state_fill = gsub("\\s\\+\\sNA$", "", state_fill)) %>%
  mutate(spec_name = paste(genus, species, author)) %>%
  select(utm_10km, period, spec_name, genus, species, author, state_fill)
```


